Sperm modifiers of female drive.

### 2013-09-12 writeup_and_notes/male_dependent_drive.tex
Yaniv's changes on Thurs

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{}
\rhead{}
\renewcommand{\headrulewidth}{0.0pt}
\rfoot{}
\cfoot{\thepage}
\date{}
\bibliographystyle{plain}
\begin{document}
\maketitle
\begin{center}
Center for Population Biology \& Department of Evolution and Ecology \\ University of California - Davis \\ Davis, CA, 95616
\end{center}

@@

%Since all alleles within an individual rely on individual survival and reproduction for evolutionary success, most of life in a diploid eukaryotic genome is harmonious.
%However, this harmony is incomplete, as in numerous cases an allele can benefit (in the short or long term) from taking advantage of its host.
Despite the apparent unity of the organism, occasionally alleles can
gain an evolutionary advantage at a cost to  individual fitness
\citep{Burt2006}, often by exploiting meiosis and gametogensis.
%A clear opportunity for this conflict arises during gametogenesis where alternative alleles are in direct competition for representation in the next generation.
Female meiosis, an asymmetric event in which only one of two alternate alleles
enters the  egg, with the others consigned to the polar body, is one such occasion \citep{Sandler1957,Pardo-ManuelDeVillena2001a}. 
%Female meiosis is particularly unfair, only one of the four products of meiosis becomes an egg, while the other threes are disgarded into polar bodies. 
An allele that biases female meiosis in its favor (i.e. a meiotic driver), may increase in frequency even if this driver entails a pleiotropic fitness cost \citep{Prout1973}, generating a genetic conflict between the success of the driver and organismal fitness.
%Alleles that can segregate into the egg more than half the time in heterozygotes, by exploiting asymmetries, 
%can potentially increase in frequency in the population (i.e. experiencing true meiotic drive).
%If such alleles have fitness consequences they are a source of genetic conflict, 
%and can become balanced in the population if their host in homozygotes 
%out weighs their ability to spread to spread through heterozygotes. 
Meiotic drivers observed in nature (in both plants
\citep{Buckler1999,Fishman2005,Fishman2008}, and animals
\citep{Agulnik1990,Wu2005,Pardo-ManuelDeVillena2001c}) highlight this
conflict -- the selfish benefits of drive and the associated
pleiotropic fitness costs sustain a balanced polymorphism
\citep{Prout1973}, 
and often generate on ongoing evolutionary escalation of drive suppressors and enhancers \citep{Dawe1996,Fishman2008}. 
%Such balanced drive systems can cause the subsequent evolution of linked enhancers of drive
%and of supressors of drive throughout the genome. 
%Indeed, the known polymorphic female meiotic drive systems (OM, maize knobs, IN, mimulus, others?) habor
%a great diversity of enhancers and supressors. 
The threat of meiotic drive to organismal fitness is potentially so
severe that it has been hypothesized that many basic properties of meiosis and o\"{o}genesis, including the initial genome doubling in meiosis I \citep{Haig1991}, the commonness of arrested female meiosis \citep{Mira1998}, and sex differences in the recombination rate \citep{Haig2010,Brandvain2012} have evolved to disrupt meiotic drive and enforce fairness.

@@

In this article we explore through simple population genetic models the consequences of alleles that influence the outcome 
of female meiosis. 
We use these models to argue that it is actually surprisingly hard for the influence of
sperm on the outcome of meiosis to drive sustained conflict. In fact we find that sperm and egg 
genomes' interests are often aligned as they are both invested in the fate of the zygote they will form (as was suggested for the In locus \citep{Pomiankowski1993}).
This suggests that females are unlikely to evolve to prevent the influence of sperm on meiosis,
and indeed features of meiosis may evolve that facilitate the interaction of sperm with female meiosis.

@@

\begin{enumerate}
\item Setup a drive-selection balanced polymorphism in std. drive model. 
Do this by imagining the sperm-influence allele arising on the background of the driver, 
so the allele has drive capabilities, and can have had time to evolve new biology. 
Evolving on the new background means that the allele suffers the fitness consequences of the 
driver.  See A in Figure \ref{Eggsperm_3_allele_cartoon} \\
\item Sperm-based enhancers of drive can't invade (can they in some situations?). \\
\item Intuition is that the driver has already driven to a frequency 
where it is held in check by its cost in homozygotes. The sperm allele 
thus can't really help as it creates zygotes which suffer the homozygous fitness consequences.
\end{enumerate}

>>>>

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{}
\rhead{}
\renewcommand{\headrulewidth}{0.0pt}
\rfoot{}
\cfoot{\thepage}
\date{}
\bibliographystyle{plain}
\begin{document}
\maketitle
\begin{center}
Center for Population Biology \& Section in Evolution and Ecology \\ University of California - Davis \\ Davis, CA, 95616
\end{center}

@@

%Since all alleles within an individual rely on individual survival and reproduction for evolutionary success, most of life in a diploid eukaryotic genome is harmonious.
%However, this harmony is incomplete, as in numerous cases an allele can benefit (in the short or long term) from taking advantage of its host.
Despite the apparent unity of the organism, occasionally alleles can gain an evolutionary advantage at a cost to  individual fitness \citep{Burt2006}.
%A clear opportunity for this conflict arises during gametogenesis where alternative alleles are in direct competition for representation in the next generation.
Female meiosis, an event in which only one of two alternate alleles enters the  egg, is one such occasion \citep{Sandler1957,Pardo-ManuelDeVillena2001a}. 
%Female meiosis is particularly unfair, only one of the four products of meiosis becomes an egg, while the other threes are disgarded into polar bodies. 
An allele that biases female meiosis in its favor (i.e. a meiotic driver), may increase in frequency even if this driver entails a pleiotropic fitness cost \citep{Prout1973}, generating a genetic conflict between the success of the driver and organismal fitness.
%Alleles that can segregate into the egg more than half the time in heterozygotes, by exploiting asymmetries, 
%can potentially increase in frequency in the population (i.e. experiencing true meiotic drive).
%If such alleles have fitness consequences they are a source of genetic conflict, 
%and can become balanced in the population if their host in homozygotes 
%out weighs their ability to spread to spread through heterozygotes. 
Meiotic drivers observed in nature (in both plants \citep{Buckler1999,Fishman2005,Fishman2008}, and animals \citep{Agulnik1990,Wu2005,Pardo-ManuelDeVillena2001c}) highlight this conflict -- the selfish benefits of drive and the associated pleiotropic fitness costs sustain a balanced polymorphism, and often generate on ongoing evolutionary escalation of drive suppressors and enhancers \citep{Dawe1996,Fishman2008}. 
%Such balanced drive systems can cause the subsequent evolution of linked enhancers of drive
%and of supressors of drive throughout the genome. 
%Indeed, the known polymorphic female meiotic drive systems (OM, maize knobs, IN, mimulus, others?) habor
%a great diversity of enhancers and supressors. 
The threat of meiotic drive to organismal fitness is potentially so severe that that many basic properties of meiosis and o\"{o}genesis, including the initial genome doubling in meiosis I \citep{Haig1991}, the commonness of arrested female meiosis \citep{Mira1998}, and sex differences in the recombination rate \citep{Haig2010,Brandvain2012} have putatively evolved to discourage the evolution of meiotic drive an enforce fairness.

@@

In this article we explore through simple models the consequences of alleles that influence the outcome 
of female meiosis. 
We use these models to argue that it is actually surprisingly hard for the influence of
sperm on the outcome of meiosis to drive sustained conflict. In fact we find that sperm and egg 
genomes' interests are often aligned as they are both invested in the fate of the zygote they will form (as was suggested for the In locus \citep{Pomiankowski1993}).
This suggests that females are unlikely to evolve to prevent the influence of sperm on meiosis,
and indeed feature of meiosis may evolve that facilitate the interaction of sperm with female meiosis.

@@

\begin{enumerate}
\item Setup a drive-selection balanced polymorphism in std. drive model. 
Do this by imagining the sperm-influence allele arising on the background of the driver, 
so the allele has drive capabilities, and can have had time to evolve new biology.

Multilocus drivers held together with inversions, may offer a relatively large mutational/functional target.

Evolving on the new background means that the allele suffers the fitness consequences of the 
driver.  See A in Figure \ref{Eggsperm_3_allele_cartoon} \\
\item Sperm-based enhancers of drive can't invade (can they in some situations?). \\
\item Intuition is that the driver has already driven to a frequency 
where it is held in check by its cost in homozygotes. The sperm allele 
thus can't really help as it creates zygotes which suffer the homozygous fitness consequences.
\end{enumerate}
